---
title: "Principal Component Analysis"
author: "Dataset life_expectancy"
output:
  html_document: default
---
This dataset contains 179 individuals and 18 variables, 3 qualitative variables are considered as illustrative.

- - -

```{r, echo = FALSE}
library(FactoMineR)
load('D:/Project/Dim Red/PCA_Variables/Workspace.RData')
```
### 1. Study of the outliers
The analysis of the graphs does not detect any outlier.

- - -

### 2. Inertia distribution
The inertia of the first dimensions shows if there are strong relationships between variables and suggests the number of dimensions that should be studied.

The first two dimensions of analyse express **62.77%** of the total dataset inertia ; that means that 62.77% of the individuals (or variables) cloud total variability is explained by the plane.
This percentage is relatively high and thus the first plane well represents the data variability.
This value is strongly greater than the reference value that equals **20.63%**, the variability explained by this plane is thus highly significant
(the reference value is the 0.95-quantile of the inertia percentages distribution obtained by simulating 6631 data tables of equivalent size on the basis of a normal distribution).

From these observations, it should be better to also interpret the dimensions greater or equal to the third one.

```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
par(mar = c(2.6, 4.1, 1.1, 2.1))
ggplot2::ggplot(cbind.data.frame(x=1:nrow(res$eig),y=res$eig[,2])) + ggplot2::aes(x=x, y=y)+ ggplot2::geom_col(fill="blue") + ggplot2::xlab("Dimension") + ggplot2::ylab("Percentage of variance") + ggplot2::ggtitle("Decomposition of the total inertia") + ggplot2::theme_light() + ggplot2::theme(plot.title = ggplot2::element_text(hjust =0.5)) + ggplot2::scale_x_continuous(breaks=1:nrow(res$eig))
```

**Figure 2 - Decomposition of the total inertia**
*The first factor is major: it expresses itself 50.59% of the data variability.*
*Note that in such a case, the variability related to the other components might be meaningless, despite of a high percentage.*

An estimation of the right number of axis to interpret suggests to restrict the analysis to the description of the first 3 axis.
These axis present an amount of inertia greater than those obtained by the 0.95-quantile of random distributions (72.57% against 29.55%).
This observation suggests that only these axis are carrying a real information.
As a consequence, the description will stand to these axis.

- - -

### 3. Description of the plane 1:2

```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
drawn <-
c("75", "33", "19", "32", "150", "147", "52")
par(mar = c(4.1, 4.1, 1.1, 2.1))
plot.PCA(res, select = drawn, axes = c(1,2), choix = 'ind', invisible = 'quali', title = '', cex = cex)
```

**Figure 3.1 - Individuals factor map (PCA)**
*The labeled individuals are those with the higher contribution to the plane construction.*

The Wilks test p-value indicates which variable factors are the best separated on the plane (i.e. which one explain the best the distance between individuals).
```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
wilks.p <-
c(Region = 1.5318748957580106e-41, Economy_status = 4.8388163006319365e-19
)
wilks.p
```
The best qualitative variable to illustrate the distance between individuals on this plane is : *Region*.
The qualitative variables *Year* cannot separate the individuals on the plane, cause they are unimodal.

```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
sample = sample(rownames(res$call$X), length(rownames(res$call$X)))
res$call$X = res$call$X[sample,]
res$ind$coord = res$ind$coord[sample[!sample %in% rownames(res$ind.sup$coord)],]
res$ind.sup$coord = res$ind.sup$coord[sample[sample %in% rownames(res$ind.sup$coord)],,drop=FALSE]
drawn <-
c("75", "33", "19", "32", "150", "147", "52")
hab <-
"Region"
par(mar = c(4.1, 4.1, 1.1, 2.1))
plotellipses(res, axes = 1:2, select = drawn, keepvar = hab, title = '', cex = cex)
```

**Figure 3.2 - Individuals factor map (PCA)**
*The labeled individuals are those with the higher contribution to the plane construction.*
*The individuals are coloured after their category for the variable* Region.

```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
drawn <-
c("Infant_deaths", "Under_five_deaths", "Polio", "Diphtheria", 
"Thinness_ten_nineteen_years", "Thinness_five_nine_years", "Schooling", 
"Hepatitis_B", "Adult_mortality", "BMI")
par(mar = c(4.1, 4.1, 1.1, 2.1))
plot.PCA(res, select = drawn, axes = 1:2, choix = 'var', title = '', cex = cex)
```

**Figure 3.3 - Variables factor map (PCA)**
*The labeled variables are those the best shown on the plane.*

```{r, echo = FALSE, fig.align = 'center', fig.height = 3.5, fig.width = 5.5}
drawn <-
c("Rest of Europe", "Africa", "European Union", "Economy_status_0", 
"Economy_status_1", "Asia", "Central America and Caribbean", 
"North America", "South America")
par(mar = c(4.1, 4.1, 1.1, 2.1))
plot.PCA(res, select = drawn, axes = 1:2, choix = 'ind', invisible = c('ind', 'ind.sup'), title = '', cex = cex)
```

**Figure 3.4 - Qualitative factor map (PCA)**
*The labeled factors are those the best shown on the plane.*

* * *

The **dimension 1** opposes individuals characterized by a strongly positive coordinate on the axis (to the right of the graph)
to individuals such as *32*, *33*, *147*, *52*, *19* and *150* (to the left of the graph, characterized by a strongly negative coordinate on the axis).

The group 1 (characterized by a positive coordinate on the axis) is sharing :

- high values for the variables *BMI*, *Schooling*, *Measles*, *Polio*, *Diphtheria*, *Hepatitis_B*, *Alcohol_consumption* and *GDP_per_capita* (variables are sorted from the strongest).
- low values for the variables *Infant_deaths*, *Under_five_deaths*, *Adult_mortality*, *Thinness_ten_nineteen_years*, *Thinness_five_nine_years* and *Incidents_HIV* (variables are sorted from the weakest).

The group in which the individuals *32*, *33*, *147* and *52* stand (characterized by a negative coordinate on the axis) is sharing :

- high values for the variables *Under_five_deaths*, *Infant_deaths*, *Adult_mortality*, *Incidents_HIV*, *Thinness_ten_nineteen_years* and *Thinness_five_nine_years* (variables are sorted from the strongest).
- low values for the variables *Polio*, *Diphtheria*, *Hepatitis_B*, *Schooling*, *BMI*, *Measles*, *GDP_per_capita* and *Alcohol_consumption* (variables are sorted from the weakest).

The group in which the individuals *19* and *150* stand (characterized by a negative coordinate on the axis) is sharing :

- high values for the variables *Thinness_five_nine_years*, *Thinness_ten_nineteen_years* and *Population_mln* (variables are sorted from the strongest).
- low values for the variables *BMI*, *Alcohol_consumption*, *Schooling*, *Measles* and *GDP_per_capita* (variables are sorted from the weakest).

Note that the variables *Africa* and *Rest of Europe* are highly correlated with this dimension (respective correlation of 0.93, 0.92). These variables could therefore summarize themselve the dimension 1.

* * *

The **dimension 2** opposes individuals such as *19* and *150* (to the top of the graph, characterized by a strongly positive coordinate on the axis)
to individuals such as *32*, *33*, *147* and *52* (to the bottom of the graph, characterized by a strongly negative coordinate on the axis).

The group in which the individuals *19* and *150* stand (characterized by a positive coordinate on the axis) is sharing :

- high values for the variables *Thinness_five_nine_years*, *Thinness_ten_nineteen_years* and *Population_mln* (variables are sorted from the strongest).
- low values for the variables *BMI*, *Alcohol_consumption*, *Schooling*, *Measles* and *GDP_per_capita* (variables are sorted from the weakest).

The group in which the individuals *32*, *33*, *147* and *52* stand (characterized by a negative coordinate on the axis) is sharing :

- high values for the variables *Under_five_deaths*, *Infant_deaths*, *Adult_mortality*, *Incidents_HIV*, *Thinness_ten_nineteen_years* and *Thinness_five_nine_years* (variables are sorted from the strongest).
- low values for the variables *Polio*, *Diphtheria*, *Hepatitis_B*, *Schooling*, *BMI*, *Measles*, *GDP_per_capita* and *Alcohol_consumption* (variables are sorted from the weakest).

